---
title: "Differential abundance analysis, semi-targeted approach"
subtitle: "Positive polarity"
author: "Christa Malfertheiner"
date: "07 November 2021"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

```{r setup, echo = FALSE, results = "asis", warning = FALSE}
library(BiocStyle)
BiocStyle::markdown()
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r parameters, echo = FALSE, warning = FALSE}
## Set general parameters
polarity <- "POS" # specify "POS" or "NEG"
p.cut <- 0.05     # cut-off for significance.
m.cut <- 0.7      # cut-off for log2 fold change
set.seed(123)
## Setting golden ratio to save images
phi <- (1+sqrt(5))/2
FILE_NAME <- "sts_targeted_pos"
## Define paths:
IMAGE_PATH <- paste0("images/", FILE_NAME, "/")
if (dir.exists(IMAGE_PATH)) unlink(IMAGE_PATH, recursive = TRUE, force = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE, showWarnings = FALSE)
RDATA_PATH <- paste0("data/RData/", FILE_NAME, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)
RESULT_PATH <- paste0("data/results/", FILE_NAME, "/")
dir.create(RESULT_PATH, recursive = TRUE, showWarnings = FALSE)
```

# Introduction

In this document we perform the differential abundance analysis of the features
from short term stability data collected with mitra tips, with the aim of 
finding significant storage-related features. This task is performed by 
hypothesis testing, where we try to identify which metabolites have the most
different concentrations.
We follow a semi-targeted approach, where we look at concentrations of features
corresponding to lab-internal set of standards.


# Data import

First, we load the required packages and the data, after preprocessing and
normalization. The end result of these steps is a `SummarizedExperiment` that
contains aligned data, where features are grouped (after correspondence), and
that have undergone gap filling, normalization by the median, linear fitting and 
per-feature between-batch normalization to remove any unwanted variability. 
The `SummarizedExperiment` lets us store all the information regarding the 
normalization steps in the form of `assays`, which we are still able to access 
to proceed with the analysis.

```{r load-data, echo = FALSE, warning = FALSE}
library(xcms)
library(limma)
library(pheatmap)
library(writexl)
library(SummarizedExperiment)
library(RColorBrewer)
library(MsFeatures)
library(CompMetaboTools)
library(pander)
load("data/RData/sts_normalization_pos/res_pos.RData")
res_pos$sample_pair <- paste0(res_pos$source, ".", res_pos$sample)
```

It is important now to remove the `POOL` samples from the dataset, because the
analysis has to be performed only on study samples; the `POOL` samples, though
are still required to evaluate the goodness of the detected features, therefore
they will be stored in a separate `SummarizedExperiment` object that can be
accessed when needed.
We also exclude the `RBC` samples from our analysis, as these samples showed
a noisy signal beforehand. We also store them separately in a 
`SummarizedExperiment` object.

We also assign the colours as seen before.

```{r split-qc, echo = TRUE}
res_qc <- res_pos[, res_pos$storage == "QC"]
res_pos <- res_pos[, res_pos$storage != "QC"]

res_pos$storage <- factor(as.character(res_pos$storage))

## Define colors for the groups.
col_storage <- brewer.pal(6, name = "Set1")
names(col_storage) <- c("4C_BAG",   # red
                        "RT",        # blue
                        "BAG",       # green
                        "VACUUM",    # purple
                        "4C_VACUUM", # orange
                        "QC")        # yellow
col_time <- brewer.pal(8, name = "Set3")
names(col_time) <- c("1d",
                     "1w",
                     "2d",
                     "2h",
                     "3d",
                     "6h",
                     "2w",
                     "QC")

## Setting golden ratio to save images
phi <- (1+sqrt(5))/2
```

The samples used in this analysis are listed below.

```{r, echo = FALSE, results = "asis"}
tab <- colData(res_pos)[, c("storage", "time")]
pandoc.table(as.data.frame(tab), style = "rmarkdown",
             caption = "Samples used in this analysis")
```


# Semi-targeted analysis

First, we restrict the analysis to a set of known compounds, whose
mass-to-charge ratio and retention time have been previously measured from the
pure standards. To start, a table of standards is loaded, then
a search in the m/z and retention time dimensions is carried out to match the
features found before to these compounds. Subsequently, EICs for all the
assigned features are plotted and visually inspected: this step is necessary to
accurately pair the detected peaks to the standards. When the list of assigned
standards is complete, we perform a differential abundance analysis on the
subset of features.

First, we load the information on the standards, with the aim of identifying
features potentially matching these and plotting the EICs for all of these.

```{r known-cmps, message = FALSE, warning = FALSE}
## Extract known compunds
library("MetaboCoreUtils")
library(Rdisop)
std_info <- read.table(
    "https://raw.githubusercontent.com/EuracBiomedicalResearch/lcms-standards/master/data/standards_dilution.txt",
    sep = "\t", header = TRUE, as.is = TRUE)
std_info <- std_info[!is.na(std_info[, "POS"]), ]
rownames(std_info) <- 1:nrow(std_info)
std_info$mzneut = NA
std_info$mz_ion = NA
for (i in seq(nrow(std_info))) {
    if (grepl("C", std_info$formula[i])) {
        std_info$mzneut[i] <- getMolecule(
            as.character(std_info$formula[i]))$exactmass
    } else {
        std_info$mzneut[i] = as.numeric(std_info$formula[i])
    }
    ## Calculate also the m/z
    std_info$mz_ion[i] <- mass2mz(
        std_info$mzneut[i], adduct = std_info[i, "POS"])[1, 1]
}
std_info <- std_info[!is.na(std_info$mz_ion), ]
std_info <- std_info[order(std_info$name), ]

dr <- paste0(IMAGE_PATH, "/standards/")
dir.create(dr, recursive = TRUE, showWarnings = FALSE)

##load data_pos
load("data/RData/sts_normalization_pos/data_pos_filled.RData")
## Subset to the samples we're currently analyzing.
tmp <- filterFile(data_pos, match(res_pos$mzML_file, data_pos$mzML_file),
                  keepFeatures = TRUE)
```

We next match features from our data set against m/z of the expected ions and
the retention times for the set of lab-internal standards, extract their ion
chromatogram and plot these.

```{r}
## Match feature's m/z and rt values against expecte values for standards.
library(MetaboAnnotation)
register(SerialParam())
rowData(res_pos)$ft <- row.names(res_pos)
## Note: would need a ppm of 40 to get Homocysteine-FT0178
par <- MzRtParam(tolerance = 0, ppm = 20, toleranceRt = 30)
mo <- matchMz(res_pos, std_info, param = par, mzColname = c("mzmed", "mz_ion"), 
              rtColname = c("rtmed", "RT"))
## Subset to matching features.
mo <- mo[whichQuery(mo)]
mo <- pruneTarget(mo)
chrs <- featureChromatograms(tmp, features = mo$ft, expandRt = 7, filled = TRUE)
sample_colors <- col_storage[tmp$storage]
for (i in seq_len(length(mo$ft))) {
    chr <- chrs[i, ]
    pks <- chromPeaks(chr)
    fl <- mo$target_name[i]
    png(paste0(dr, fl, "-", mo$ft[i], ".png"),
        width = 10, height = 8, units = "cm", res = 300, pointsize = 6)
    plot(chr, col = "#00000040",
         peakCol = paste0(sample_colors[pks[, "column"]], 50),
         peakBg = paste0(sample_colors[pks[, "column"]], 10))
    abline(v = mo$target_RT[i])
    legend("topleft", legend = c(mo$ft[i], fl,
                                 paste0("rt: ", mo$target_RT[i]),
                                 paste0("mz: ", mo$target_mz_ion[i])))
    dev.off()
}
```

The EICs have been manually inspected and the best matching feature has been
manually assigned to the corresponding standard.

```{r assign-feature-metabolite, echo = FALSE, warning = FALSE, message = FALSE}
## This is the tricky manual thing:
## - Go through all plots for all standards and if there is one feature
##   that clearly matches (i.e. retention time close to the expected retention
##   time and a single peak present in the wider rt range) assign it.
to_keep <- c(FT00602 = "1-Methylhistidine",
             FT00603 = "3-Methylhistidine", ## same as 1-Methylhistidine
             FT00320 = "5-Oxoproline",
             FT01648 = "8-Oxo-2-Deoxyguanosine",
             FT00912 = "Acetylcarnitine",
             FT01254 = "Acetyl-Glucosamine",
             FT00850 = "Acetylhistidine",
             FT00901 = "ADMA",
             FT03607 = "ADP",
             FT00127 = "Alanine",
             FT00546 = "alpha-Aminoadipic acid",
             FT02791 = "alpha-Lactose", 
             FT02548 = "AMP",
             FT00643 = "Arginine",
             FT00338 = "Asparagine",
             FT00239 = "Betaine",
             #FT01164 = "C4 Carnitine",
             #FT01276 = "C5 Carnitine", #looks noisy
             FT00824 = "Caffeine",
             FT01117 = "Carnosine",
             FT00167 = "Choline",
             FT00655 = "Citrulline",
             ## c("Corticosterone", "FT02541"), # rt shifted 20s
             ##c("Creatine", "FT00334"),
             FT00219 = "Creatinine",
             #FT01973 = "dCMP",
             FT03607 = "dGDP",
             FT02549 = "dGMP",
             FT00894 = "Fructose", # same as glucose and mannose
             FT00917 = "Galactitol", # same as sorbitol
             FT00894 = "Glucose", # same as Fructose
             FT00444 = "Glutamine",
             FT06228 = "Glutathione Oxidized", # same as Glu. Reduced
             FT06228 = "Glutathione Reduced", # same as Glu. Oxidized
             FT01379 = "Glycero-phosphocholine",
             FT00060 = "Glycine",
             FT01648 = "Guanosine",
             FT00204 = "Histamine",
             FT00508 = "Histidine",
             FT00333 = "Hydroxyproline",
             FT00374 = "Hypoxanthine",
             FT00335 = "Isoleucine",
             FT00980 = "L,L-Cyclo(leucylprolyl)",
             FT00348 = "L-Aspartic Acid", 
             FT00554 = "L-Carnitine", 
             FT00253 = "L-Cysteine",
             FT00335 = "Leucine", #same as Isoleucine
             FT00457 = "L-Glutamic Acid",
             FT00447 = "Lysine",
             FT00894 = "Mannose", # same as Fructose and Glucose
             FT00469 = "Methionine",
             FT00575 = "Methioninesulfoxide", 
             ## c("Myoinositol", "FT00894"), # same as fructose, glucose and mannose
             FT00641 = "N-Acetylornithine",
             FT06938 = "NAD",
             FT00272 = "Niacinamide",
             FT00340 = "Ornithine",
             FT03267 = "Palmitoylcarnitine", 
             FT00578 = "Phenylalanine",
             FT00255 = "Phenylethylamine",
             FT00721 = "Phosphorylcholine",
             FT00414 = "Phosphorylethanolamine",
             ##c("Pipecolic acid", "FT00322"),
             FT00232 = "Proline",
             FT03079 = "SAH",
             FT03263 = "SAMe",
             FT00901 = "SDMA",
             FT00179 = "Serine",
             FT00917 = "Sorbitol", 
             FT03017 = "Sphingosine-1-phosphate",
             FT01866 = "Sphingosine",
             FT02791 = "Sucrose", 
             FT00295 = "Taurine",
             FT00245 = "Threonine",
             FT00920 = "Tryptophan",
             FT00695 = "Tyrosine",
             FT00239 = "Valine")
mo <- filterMatches(mo, queryValue = names(to_keep), 
                    targetValue = to_keep, 
                    queryColname = "ft",
                    targetColname = "target_name",
                    keep = TRUE)
mo <- mo[whichQuery(mo)]
mo <- pruneTarget(mo) # Probably it's not very useful but since these two lines
# are used one after the other we could create a function e.g. reduceMatched 
# that does the two things
## Handling duplicates.
md <- as.data.frame(matchedData(mo, c("ft", "target_name", "target_HMDB.code")))
md <- split(md, md$ft)
md <- do.call(rbind, lapply(md, function(z) {
    tmp <- data.frame(ft = z$ft[1L])
    tmp$name <- paste0(z$target_name, collapse = ";")
    tmp$HMDB <- paste0(z$target_HMDB.code, collapse = ";")
    tmp
}))
```

A total of `r sum(nrow(md))` standards have been identified. The features
identified and the corresponding metabolite are summarized in this table:

```{r result-table-ft-std, echo = FALSE, results = "asis"}
## Write result table
md <- md[order(md$name), ]
pandoc.table(md[, c("ft", "name")], style = "rmarkdown",
             caption = "Features assigned to known compounds")
```

Next, only the features assigned to the standards are taken into consideration
and are subsetted in the `std_res` object.

```{r std-subset, echo = FALSE}
std_res <- query(mo)
rowData(std_res) <- cbind(rowData(std_res),
                          md[rownames(std_res), c("name", "HMDB")])
```

The subsetting reduced the number of features to `r length(std_res)`. 

A PCA analysis is then performed on the subset to verify whether anything has
changed and if any similarities among the samples are visible or not.

```{r standards-pca-all, echo = FALSE}
pc <- prcomp(t(log2(assay(std_res, "normalized_filled_imputed"))),
                 center = TRUE, scale. = FALSE)
```

```{r standards-pca-plot, fig.path = IMAGE_PATH, fig.cap = "PCA of the samples based on intensities of known compounds.", fig.width = 7 * phi, fig.height = 7, echo = FALSE}
par(mfrow = c(1, 2))
plot_pca(pc, col = paste0(col_storage[as.character(std_res$storage)], 90),
         pc_x = 1, pc_y = 2, labels = std_res$differentiation)
plot_pca(pc, col = paste0(col_storage[as.character(std_res$storage)], 90),
         pc_x = 3, pc_y = 4)
legend("topleft", col = col_storage, legend = names(col_storage),
       title = "phenotype", pch = 16, ncol = 2)
```

In the PC1 plot, we see a separation of samples stored at room temperture (RT),
samples that were stored under other circumstances cannot be distinguished from
the PCA plots.

## Heatmap 

To visualize the intensity of all detected standards across the three matrices,
we created a heatmap of the known compounds we found in our samples.

```{r heatmap-source-semitargeted, fig.width = 15, fig.height = 13, fig.cap = "Heatmap of known compounds. Note that for better visibility, the color range has been restricted to -5 to 5, thus, differences larger than these values are assigned the extreme colors.", echo = FALSE, fig.path = IMAGE_PATH}
## Create heatmap
tmp <- log2(assay(
    std_res, "normalized_filled_imputed")[rownames(std_res), , drop = FALSE])
tmp <- tmp - rowMeans(tmp, na.rm = TRUE)
rownames(tmp) <- rowData(std_res)$name
colnames(tmp) <- std_res$mzML_file
ann <- as.data.frame(colData(std_res)[, c("storage", "time")])
rownames(ann) <- colnames(tmp)
pm <- pheatmap(tmp, annotation_col = ann, labels_col = colnames(tmp),
               breaks = seq(-5, 5, length.out = 101),
               annotation_color = list(storage = col_storage,
                                   time = col_time))
```

In the heatmap we see two main cluster, both containing samples from all
storage conditions. The second cluster contains all samples that were stored
for 6 hours and most of the samples stored for one day, whereas the first 
cluster contains most of the smples stored for two days.

The metabolites also cluster into different group. One cluster contains only
**Sphingosine-1-phosphate** (FT03017) which shows a low intensity in samples
stored for 6h or 1d, but has a high intensity all other samples. In contrast, 
another cluster of **alpha-Lactose, Sucrose and SAMe** shows higher 
concentrations in the second cluster compared to the first.

Now, we start with the differential abundance analysis.

## Differential abundance analysis

We next identify metabolites with significant differences in their abundances
between the sample matrices. The differential abundance analysis is performed
on the subset of features that have previously been assigned to the
standards. We apply feature-wise multiple linear regression using the `lmFit`
function and we add the matrix defining the contrasts using
`contrast.fit`. Then, we calculate the p-values with `eBayes`.  Subsequently, we
generate a data frame with the coefficients, the raw and adjusted p-values (we
apply a Benjamini-Hochberg correction for better control of the false discovery
rate), the average intensity of signals in plasma samples, capillary and venous
blood samples and whether or not a feature is to be considered significant.


```{r analysis}
## Factor sample source, sex and age
storage <- factor(std_res$storage)
time <- factor(std_res$time)
storetime <- factor(std_res$storage_time)

## Fit the data to the desired design
dsgn <- model.matrix(~ 0 + storetime)
fit <- lmFit(log2(assay(std_res, "normalized_filled_imputed")), design = dsgn)

## Fit the actual contrasts of interest
contr_mat <- makeContrasts(
  RT2hvsRT6h = storetimeRT_2h - storetimeRT_6h,
  RT2hvsRT1d = storetimeRT_2h - storetimeRT_1d,
  RT2hvsRT2d = storetimeRT_2h - storetimeRT_2d,
  RT2hvsRT3d = storetimeRT_2h - storetimeRT_3d,
  RT2hvsRT1w = storetimeRT_2h - storetimeRT_1w,
  RT2hvsRT2w = storetimeRT_2h - storetimeRT_2w,
  RT2hvsBAG6h = storetimeRT_2h - storetimeBAG_6h,
  RT2hvsBAG1d = storetimeRT_2h - storetimeBAG_1d,
  RT2hvsBAG2d = storetimeRT_2h - storetimeBAG_2d,
  RT2hvsBAG3d = storetimeRT_2h - storetimeBAG_3d,
  RT2hvsBAG1w = storetimeRT_2h - storetimeBAG_1w,
  RT2hvsBAG2w = storetimeRT_2h - storetimeBAG_2w,
  RT2hvsVACUUM6h = storetimeRT_2h - storetimeVACUUM_6h,
  RT2hvsVACUUM1d = storetimeRT_2h - storetimeVACUUM_1d,
  RT2hvsVACUUM2d = storetimeRT_2h - storetimeVACUUM_2d,
  RT2hvsVACUUM3d = storetimeRT_2h - storetimeVACUUM_3d,
  RT2hvsVACUUM1w = storetimeRT_2h - storetimeVACUUM_1w,
  RT2hvsVACUUM2w = storetimeRT_2h - storetimeVACUUM_2w,
  RT2hvs4CBAG6h = storetimeRT_2h - storetime4C_BAG_6h,
  RT2hvs4CBAG1d = storetimeRT_2h - storetime4C_BAG_1d,
  RT2hvs4CBAG2d = storetimeRT_2h - storetime4C_BAG_2d,
  RT2hvs4CBAG3d = storetimeRT_2h - storetime4C_BAG_3d,
  RT2hvs4CBAG1w = storetimeRT_2h - storetime4C_BAG_1w,
  RT2hvs4CBAG2w = storetimeRT_2h - storetime4C_BAG_2w,
  RT2hvs4CVACUUM6h = storetimeRT_2h - storetime4C_VACUUM_6h,
  RT2hvs4CVACUUM1d = storetimeRT_2h - storetime4C_VACUUM_1d,
  RT2hvs4CVACUUM2d = storetimeRT_2h - storetime4C_VACUUM_2d,
  RT2hvs4CVACUUM3d = storetimeRT_2h - storetime4C_VACUUM_3d,
  RT2hvs4CVACUUM1w = storetimeRT_2h - storetime4C_VACUUM_1w,
  RT2hvs4CVACUUM2w = storetimeRT_2h - storetime4C_VACUUM_2w,
  levels = dsgn)
fit <- contrasts.fit(fit, contrasts = contr_mat)
fit <- eBayes(fit)
adjp <- apply(fit$p.value, 2, p.adjust, method = "BH")
tmp <- data.frame(
    coef = fit$coefficient,
    pvalue = fit$p.value,
    adjp = adjp,
    significant = adjp < p.cut & abs(fit$coefficient) > m.cut
)
tmp$avg.RT2h <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_2h"]))
tmp$avg.RT6h <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_6h"]))
tmp$avg.RT1d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_1d"]))
tmp$avg.RT2d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_2d"]))
tmp$avg.RT3d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_3d"]))
tmp$avg.RT1w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_1w"]))
tmp$avg.RT2w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "RT_2w"]))
tmp$avg.BAG6h <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "BAG_6h"]))
tmp$avg.BAG1d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "BAG_1d"]))
tmp$avg.BAG2d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "BAG_2d"]))
tmp$avg.BAG3d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "BAG_3d"]))
tmp$avg.BAG1w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "BAG_1w"]))
tmp$avg.BAG2w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "BAG_2w"]))
tmp$avg.VACUUM6h <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "VACUUM_6h"]))
tmp$avg.VACUUM1d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "VACUUM_1d"]))
tmp$avg.VACUUM2d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "VACUUM_2d"]))
tmp$avg.VACUUM3d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "VACUUM_3d"]))
tmp$avg.VACUUM1w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "VACUUM_1w"]))
tmp$avg.VACUUM2w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "VACUUM_2w"]))
tmp$avg.4CBAG6h <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_BAG_6h"]))
tmp$avg.4CBAG1d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_BAG_1d"]))
tmp$avg.4CBAG2d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_BAG_2d"]))
tmp$avg.4CBAG3d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_BAG_3d"]))
tmp$avg.4CBAG1w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_BAG_1w"]))
tmp$avg.4CBAG2w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_BAG_2w"]))
tmp$avg.4CVACUUM6h <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_VACUUM_6h"]))
tmp$avg.4CVACUUM1d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_VACUUM_1d"]))
tmp$avg.4CVACUUM2d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_VACUUM_2d"]))
tmp$avg.4CVACUUM3d <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_VACUUM_3d"]))
tmp$avg.4CVACUUM1w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_VACUUM_1w"]))
tmp$avg.4CVACUUM2w <- rowMeans(log2(assay(std_res, "normalized_filled_imputed")
                              [, std_res$storage_time == "4C_VACUUM_2w"]))
rowData(std_res) <- cbind(rowData(std_res), tmp)
```

We plot then the distribution of p-values, both raw and adjusted for all 
samples stored at room temperature:

```{r standards-p-value-histogram, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 7 * phi, fig.cap = "Distribution of raw (left) and adjusted p-values (right) for all 6 comparisons."}
par(mfrow = c(3, 4))
hist(rowData(std_res)$pvalue.RT2hvsRT6h, breaks = 64, xlab = "p value",
     main = "RT for 2h vs RT for 6h")
hist(rowData(std_res)$adjp.RT2hvsRT6h, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "RT for 2h vs RT for 6h")
hist(rowData(std_res)$pvalue.RT2hvsRT1d, breaks = 64, xlab = "p value",
     main = "RT for 2h vs RT for 1d")
hist(rowData(std_res)$adjp.RT2hvsRT1d, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "RT for 2h vs RT for 1d")
hist(rowData(std_res)$pvalue.RT2hvsRT2d, breaks = 64, xlab = "p value",
     main = "RT for 2h vs RT for 2d")
hist(rowData(std_res)$adjp.RT2hvsRT2d, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "RT for 2h vs RT for 2d")
hist(rowData(std_res)$pvalue.RT2hvsRT3d, breaks = 64, xlab = "p value",
     main = "RT for 2h vs RT for 3d")
hist(rowData(std_res)$adjp.RT2hvsRT3d, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "RT for 2h vs RT for 3d")
hist(rowData(std_res)$pvalue.RT2hvsRT1w, breaks = 64, xlab = "p value",
     main = "RT for 2h vs RT for 1w")
hist(rowData(std_res)$adjp.RT2hvsRT1w, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "RT for 2h vs RT for 1w")
hist(rowData(std_res)$pvalue.RT2hvsRT2w, breaks = 64, xlab = "p value",
     main = "RT for 2h vs RT for 2w")
hist(rowData(std_res)$adjp.RT2hvsRT2w, breaks = 64,
     xlab = expression(p[BH]~value),
     main = "RT for 2h vs RT for 2w")

```

Thus, most of the metabolites were identified to have significantly different
concentrations between the comparison. A table with the number of significant
metabolites is shown below.

```{r table-sig, echo = FALSE, results = "asis"}
tab <- colSums(as.matrix(rowData(std_res)[, grep("significant", 
                                             colnames(rowData(std_res)))]))
pandoc.table(tab, style = "rmarkdown",
             caption = paste0("Number of significant features of the in",
                              " total", nrow(std_res), "analyzed features."))
```

The number of significant features is then shown in a barplot:

```{r sig-features-barplot, echo = FALSE, fig.path = IMAGE_PATH, fig.width = 10 * phi, fig.cap = "Amount of significant features per storage condition over time"}

pd <- read.table("data/sts_semi_sig.txt", sep = "\t", as.is = TRUE, header = TRUE)
library("ggplot2")
ggplot(pd, aes(fill=storage, y=significant, x=time)) + 
    geom_bar(position="dodge", stat="identity") +
  scale_fill_brewer(palette="Set1")
```
Samples that were stored at room temperature show the least change in the 
beginning of storage duration, but after one and two weeks, those samples
have the highest number of significant features when compared to the samples
that were freezed directly after drying. 

# Session information

```{r}
sessionInfo()
```

